\textbf{Appendix}\\ A Thee-Step Approach for Assessing Landscape Connectivity via Simulated Dispersal: African Wild Dog Case Study

### 2022-01-24 Chapter_1/04_Manuscript/Manuscript.tex
updates

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{A Three Steps Workflow for Simulating Dispersal Trajectories and
Assessing Landscape Connectivity: A Case Study on the Endangered African Wild
Dog}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
% \newpage
% \begin{abstract}
% %------------------------------------------------------------------------------
% % Short Version
% %------------------------------------------------------------------------------
% \textbf{Short Version }
% The ability to disperse is contingent on a sufficient degree of landscape
% connectivity, which is why the identification and preservation of movement
% corridors that promote connectivity has become a task of extraordinary
% importance. Currently, ecologists rely on least-cost analysis or circuit theory
% to investigate connectivity, although both methods make several assumptions that
% are hardly met in reality. To address these issues, simulations from
% individual-based movement models have been proposed, yet a unified framework to
% simulate dispersal and quantify connectivity is lacking.
%
% Here, we propose a simple three-step workflow that combines several
% already-existing methods to assess connectivity using simulated dispersal
% trajectories. In step one, we use integrated step selection functions to
% parametrize a mechanistic movement model rendering dispersal behavior. In step
% two, we apply the parametrized model as an individual-based movement model to
% simulate dispersal trajectories. In step three, we combine simulated
% trajectories into three complementary connectivity maps, each focusing on a
% different aspect of landscape connectivity.
%
% We showcase the application of the proposed workflow using empirical data of
% dispersing African wild dogs (\textit{Lycaon pictus}) and assess landscape
% connectivity within the world's largest transboundary conservation area in
% Southern Africa. We thereby shed light into dispersing wild dogs' habitat and
% movement preferences, while also uncovering crucial dispersal corridors. With
% this analysis, we demonstrate that simulations from integrated step selection
% functions offer a simple, yet powerful alternative to traditional connectivity
% modeling techniques.
% \end{abstract}

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
Dispersal of individuals contributes to long-term population persistence, yet
requires a sufficient degree of landscape connectivity. To date, connectivity
has mainly been investigated using least-cost analysis and circuit theory, two
methods that make assumptions that are hardly applicable during dispersal.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route to reach it. Circuit theory
presumes a complete random walk and fails to incorporate directional
persistence. While these assumptions can be relaxed by simulating dispersal
movements spatio-temporally explicitly across the landscape of interest, a
unified approach for such simulations is lacking.

Here, we present a simple three-step workflow to simulate dispersal movements
and to assess connectivity starting from empirical GPS movement data. In step
one, we use integrated step selection functions to fit a mechanistic movement
model describing habitat and movement preferences of dispersing individuals. In
step two, we apply the parameterized movement model to simulate dispersal
trajectories. In step three, we derive three complementary connectivity maps: a
heatmap that highlights frequently traversed areas, a betweenness map that
pinpoints dispersal corridors, and a map of inter-patch connectivity that
indicates the presence and intensity of functional links between habitat
patches. As a case study, we applied the workflow to GPS data collected during
dispersal of the endangered African wild dog (\textit{Lycaon pictus}) inhabiting
the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA).

Results from the case study show that wild dogs preferably disperse with
directional persistence in the vicinity to water bodies, through areas with
little human influence and sparse woodland cover. Dispersal simulations and
subsequent connectivity maps revealed several dispersal hotspots and corridors
across the extent of the KAZA-TFCA. Connectivity between NPs inside the
KAZA-TFCA was good, yet few dispersers successfully moved from Zambia's NPs into
neighboring areas. By rendering interactions between movement preferences and
habitat preferences, our movement model substantially outperformeed a previous
model that omitted such interactions. While we simulated 80,000 dispersal
trajectories in total, we found that metrics of connectivity only marginally
changed beyond 10,500 simulations, suggesting that fewer simulations would have
sufficed.

Ultimately, we show that a simulation-based approach that leverages on
step-selection functions offers a simple yet powerful alternative to traditional
connectivity modeling techniques. Such an approach not only makes fewer
biologically unrealistic assumptions but also permits a more mechanistic
understanding of dispersal and landscape connectivity. Our workflow is thus
useful for a variety of applications in ecological, evolutionary, and
conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. Importantly, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal movements, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used models to assess connectivity are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}). Unfortunately, both models rest on assumptions that appear
unsuitable for dispersers, calling for the development of alternative
approaches. One promising alternative to investigate dispersal and assess
connectivity is to simulate dispersal from individual-based movement models
\citep{Diniz.2019}.

@@

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
from individual-based movement models (IBMMs) and by converting simulated
trajectories into meaningful measures of connectivity \citep{Diniz.2019}). In
contrast to LCPA and CT, IBMMs allow to explicitly simulate how individuals move
across and interact with the encountered landscape \citep{Kanagaraj.2013,
Clark.2015, Allen.2016, Hauenstein.2019, Zeller.2020}, as well as to render
potential interactions between movement behavior and habitat conditions
\citep{Avgar.2016}. This strictly shifts the focus from a structural to a more
functional view on connectivity \citep{Tischendorf.2000, Kanagaraj.2013,
Hauenstein.2019}. Furthermore, IBMMs generate movement sequentially, i.e. they
generate a series of steps, the temporal dimension of dispersal movements
becomes explicit and allows to model autocorrelation between consecutive steps
\citep{Diniz.2019}. Finally, simulations from IBMMs do not enforce movement or
connections towards preconceived endpoints, thereby preventing biases arising
from misplaced endpoints. Despite these advantages, a unified framework to
simulate dispersal and assess connectivity using IBMMs is lacking.

% Proposed Solution: Three-Step Workflow
\subsection{Proposed Solution: Three-Step Workflow}
Here, we propose a simple three-step workflow for simulating dispersal and
assessing landscape connectivity (\Cref{GraphicalAbstract}). In step one, we
combine GPS movement data of dispersing individuals with relevant habitat
covariates to fit a mechanistic movement model using integrated step selection
functions (ISSFs, \citealp{Avgar.2016}). ISSFs allow inference on the study
species' habitat kernel (i.e. habitat preferences), its movement kernel (i.e.
movement preferences/capabilities), and potential interactions between the two
\citep{Avgar.2016, Fieberg.2021}. In step two, we simulate individual dispersal
trajectories using the parametrized movement model. Comparable simulations have
already been applied to estimate steady-state utilization distributions of
resident individuals \citep{Potts.2013, Signer.2017} and to model landscape
connectivity, yet disregarding interdependencies between habitat and movement
kernels \citep{Clark.2015, Zeller.2020}. Finally, in step three, we convert
simulated trajectories into three complementary connectivity maps, each
highlighting a different aspect of connectivity. The set of maps includes (i) a
heatmap revealing areas that are frequently traversed by dispersers (e.g.
\citealp{Hauenstein.2019, Zeller.2020}), (ii) a betweenness-map delineating
dispersal corridors and bottlenecks (e.g. \citealp{BastilleRousseau.2018}),
(iii) and a map of inter-patch connectivity, depicting the presence and
intensity of specific connections, as well as the average dispersal duration
required to connect corresponding habitat patches (e.g. \citealp{Gustafson.1996,
Kanagaraj.2013}).

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed workflow (\Cref{GraphicalAbstract})
using GPS movement data collected on dispersing individuals of the endangered
African wild dog (\textit{Lycaon Pictus}). The African wild dog is a highly
mobile species whose population persistence heavily relies on the availability
of large, natural or semi-natural landscapes and a sufficient degree of
connectivity among remaining subpopulations. Once common throughout sub-Saharan
Africa, this species has disappeared from much of its historic range, largely
due to human persecution, habitat fragmentation, and disease outbreaks
\citep{Woodroffe.2012}. Wild dogs typically disperse in single-sex coalitions
\citep{McNutt.1996, Behr.2020} and are capable of dispersing several hundred
kilometers \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020}. Although
previous studies have investigated connectivity using LCPA \citep{Hofmann.2021}
or CT \citep{Brennan.2020}, a more comprehensive and mechanistic understanding
of connectivity is missing for this species (but see \citealp{Creel.2020}). With
fewer than 6,000 free-ranging wild dogs remaining in fragmented subpopulations
\citep{Woodroffe.2012}, reliable information on landscape connectivity is
essential for the conservation of this endangered carnivore. Here, we use GPS
data of 16 dispersing wild dogs originating from a free-ranging population in
northern Botswana to parametrize a mechanistic movement model, which we then
employed to simulate 80,000 dispersal trajectories across the landscape of the
world's largest transboundary conservation area, the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA). We anticipated that simulations
based on our three-step workflow would overcome several of the above highlighted
conceptual shortcomings of traditional connectivity models and provide a more
comprehensive view on dispersal behavior and landscape connectivity.

@@

\section{Methods}
\subsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA),
an area of approximately 1.3 Mio. km\textsuperscript{2} (\Cref{StudyArea}a and
b). The KAZA-TFCA is the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia, thus hosting
a rich diversity of landscapes, ranging from savannah to grassland and from dry
to moist woodland habitats. In its center lies the Okavango Delta, a dominant
hydro-geographical feature and the world's largest flood-pulsing inland delta.
Large portions of the KAZA-TFCA are formally protected in the form of national
parks (NPs) or other protected areas, yet considerable human-dominated
landscapes (e.g. roads, agricultural sites, and settlements) remain.

@@

\subsection{Data Collection and Preparation}
\subsubsection{GPS Data}
We collected GPS movement data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions) between 2011 and 2019 from a free-ranging
population in northern Botswana (details on the data collection can be found in
\cite{Cozzi.2020} and \cite{Hofmann.2021}). To delineate periods of dispersal,
we determined the exact timing of emigration from the natal pack and settlement
in a new territory using direct field observations and through visual inspection
of the net squared displacement (NSD) metric. The NSD metric measures the
squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which we set to the center of the territory of each
dispersers' natal pack. Thus, dispersal was deemed to have started once an
individual left its natal territory and ended once the NSD metric remained
constant, indicating settlement. Because behavior during dispersal is more
pertinent to landscape connectivity than behavior during residence
\citep{Elliot.2014, Abrahms.2017}, we only considered data collected during
dispersal for our analysis. During dispersal, GPS collars recorded a fix every 4
hours and regularly transmitted data over the Iridium satellite system. To
ensure regular time intervals between GPS fixes, we removed any fixes that were
not successfully obtained at the desired 4-hour schedule (allowing for a
tolerance of \( \pm \) 15 minutes). We then converted the fixes (n = 4'169) into
steps, where each step represented the straight-line movement between two
consecutive GPS fixes \citep{Turchin.1998}.

@@

The k-fold cross-validation of the movement model showed that the final model
substantially outperformed a random guess and suggested reliable predictions
(confidence intervals of \(\bar{r}_{s, realized}\) and \(\bar{r}_{s, random}\)
do not overlap). Moreover, the model correctly assigned high selection scores to
realized steps (\Cref{MovementModel}b), indicating a good fit between
predictions and observations. Compared to the base model (\(\bar{r}_{s,
realized} = -0.55; 95\%-CI = [-0.57, -0.52]\); \citealp{Hofmann.2021}), the
inclusion of several interactions between movement and habitat covariates
significantly improved model performance (\(\bar{r}_{s, realized} = -0.65;
95\%-CI = [-0.67, -0.64]\).

@@

\begin{figure}
  \includegraphics[width=\textwidth]{99_InterpatchConnectivity.png}
  \caption{Map of inter-patch connectivity, highlighting connections between NPs
  (dark green). Yellow bubbles represent the center of the different NPs and are
  sized in relation to the number of simulated dispersers originating from each
  park. Black dots represent NPs that were smaller than 700
  km\textsuperscript{2} and therefore did not serve as source areas. Arrows
  between NPs illustrate between which NPs the simulated dispersers successfully
  moved and the color of each arrow shows the average number of steps (4-hourly
  movements) that were necessary to realize those connections. Additionally, the
  line thickness indicates the relative number of dispersers originating from a
  NP that realized those connections. Note that a similar network view could be
  adopted to investigate connectivity between other protected areas need not to
  be restricted to NPs.}
  \label{InterpatchConnectivity}
\end{figure}

@@

% Short Summary
\subsection{Short Summary}
We used empirical data from a free-ranging population of African wild dogs to
demonstrate the application of a three-step workflow to simulate dispersal
trajectories and assess landscape connectivity. In step one, we used ISSFs to
parametrize a fully mechanistic movement model describing how dispersing wild
dogs move through the landscape. In step two, we employed the movement model to
simulate 80,000 dispersal trajectories across the extent of the KAZA-TFCA, the
world's largest transboundary conservation area. In step three, we translated
the simulated trajectories into three complementary connectivity maps, each
emphasizing a different aspect of landscape connectivity (e.g. frequently
traversed areas, critical dispersal corridors and bottlenecks, and the presence
and intensity of functional links between suitable patches). Importantly, such
simulations from ISSFs overcome several conceptual shortcomings inherent to more
traditional connectivity modeling techniques, such as LCPA and CT.

% Movement Model
\subsection{Movement Model}
Our results on habitat preferences showed that dispersers avoid areas dominated
by humans and covered by water, but select for regions with open grassland in
the vicinity to water bodies. This largely complied with previous studies that
investigated habitat selection by dispersing wild dogs
\citep{DaviesMostert.2012, Masenga.2016, Woodroffe.2019, Oneill.2020,
Hofmann.2021}. However, by also accounting for movement preferences, we were
able to model several additional complexities common to dispersal. For instance,
by including an interaction between turning angle and step length we could
accommodate that dispersers exhibit step lengths that are correlated with
turning angles, meaning turning angles are smaller when individuals move fast.
Although similar autocorrelations could be incorporated by sampling step lengths
and turning angles from copula probability distributions \citep{Hodel.2021a,
Hodel.2021b}, the ISSF framework allowed us to conventiently include
correlations in the movement model. We only considered first order
autocorrelation, i.e. correlation between two consecutive steps, although higher
order autocorrelation is conceivable and may be desirable to model
\citep{Dray.2010, McClintock.2012}. This will, however, require vast amounts of
GPS data that are not interrupted by missing fixes; something that is rarely
achieved in reality \citep{Graves.2006}. The power and flexibility of ISSFs to
model additive effects between habitat and movement covariates
\citep{Avgar.2016, Signer.2021} furthermore allowed us to formally capture that
dispersing wild dogs move slower and more tortuous in areas covered by water,
something that surely was to be expected. Overall, the inclusion of interactions
between habitat and movement covariates in our movement model lead to a
significant improvement in predictive performance compared to an earlier model
that omitted such interactions \citep{Hofmann.2021}.

\subsection{Simulation}
Our simulation of 80,000 dispersal trajectories moving 2'000 steps across the
KAZA-TFCA required five days of computation on a desktop machine (AMD Ryzen 7
2700X processor with 8 x 3.6 GHz and 16 logical cores, 64 GB of RAM). The long
simulation time was primarily caused by the massive extent of the study area
considered (ca. 1.8 Mio km\textsuperscript{2}) and the large number of
trajectories simulated. Most connectivity studies focus on smaller study areas
(e.g. \citealp{Kanagaraj.2013, Clark.2015, McClure.2016, Abrahms.2017,
Zeller.2020}) and will therefore achieve faster simulation times (given the same
spatial resolution across covariates). We also believe that fewer simulated
trajectories will often suffice, as the relative traversal frequency by
simulated trajectories through randomly placed checkpoints across our study area
converged already after 10,500 runs. The number of required simulations to
achieve reliable estimates of connectivity will, however, vary depending on the
structure of the landscape and the dispersal behavior of the focal species
\citep{Gustafson.1996}. For species that disperse short distances through
homogeneous environments, few simulations may suffice to gauge connectivity,
whereas for species that disperse over long distances through heterogeneous
habitats, a large number of simulations will be required to sufficiently explore
the spectrum of possible routes.

>>>>

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{A Thee-Step Approach for Assessing Landscape Connectivity via Simulated
Dispersal: African Wild Dog Case Study}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
% \newpage
% \begin{abstract}
% %------------------------------------------------------------------------------
% % Short Version
% %------------------------------------------------------------------------------
% \textbf{Short Version }
% The ability to disperse is contingent on a sufficient degree of landscape
% connectivity, which is why the identification and preservation of movement
% corridors that promote connectivity has become a task of extraordinary
% importance. Currently, ecologists rely on least-cost analysis or circuit theory
% to investigate connectivity, although both methods make several assumptions that
% are hardly met in reality. To address these issues, simulations from
% individual-based movement models have been proposed, yet a unified framework to
% simulate dispersal and quantify connectivity is lacking.
%
% Here, we propose a simple three-step approach that combines several
% already-existing methods to assess connectivity using simulated dispersal
% trajectories. In step one, we use integrated step selection functions to
% parametrize a mechanistic movement model rendering dispersal behavior. In step
% two, we apply the parametrized model as an individual-based movement model to
% simulate dispersal trajectories. In step three, we combine simulated
% trajectories into three complementary connectivity maps, each focusing on a
% different aspect of landscape connectivity.
%
% We showcase the application of the proposed approach using empirical data of
% dispersing African wild dogs (\textit{Lycaon pictus}) and assess landscape
% connectivity within the world's largest transboundary conservation area in
% Southern Africa. We thereby shed light into dispersing wild dogs' habitat and
% movement preferences, while also uncovering crucial dispersal corridors. With
% this analysis, we demonstrate that simulations from integrated step selection
% functions offer a simple, yet powerful alternative to traditional connectivity
% modeling techniques.
% \end{abstract}

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
Dispersal of individuals contributes to long-term population persistence, yet
requires a sufficient degree of landscape connectivity. To date, connectivity
has mainly been investigated using least-cost analysis and circuit theory, two
methods that make assumptions that are hardly applicable to dispersal.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route. Circuit theory presumes a complete
random walk and fails to incorporate directional persistence. While these
assumptions can be relaxed by simulating dispersal movements explicitly across
the landscape of interest, a unified approach for such simulations is lacking.

Here, we present a simple three-step approach to simulate dispersal movements
and to assess connectivity using empirical GPS movement data and environmental
habitat layers. In step one, we use integrated step selection functions to fit a
mechanistic movement model describing habitat and movement preferences of
dispersing individuals. In step two, we apply the parameterized movement model
to simulate dispersal trajectories. In step three, we derive three complementary
connectivity maps: a heatmap that highlights frequently traversed areas, a
betweenness map that pinpoints dispersal corridors, and a map of inter-patch
connectivity that indicates the presence and intensity of functional links
between habitat patches. As a case study, we applied the approach to GPS data
from dispersing individuals of the endangered African wild dog (\textit{Lycaon
pictus}) inhabiting the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA).

Results from the case study show that wild dogs preferably disperse with
directional persistence in the vicinity to water bodies, through areas with
little human influence and sparse woodland cover. Dispersal simulations and
subsequent connectivity maps revealed several dispersal hotspots and corridors
across the extent of the KAZA-TFCA. Connectivity between NPs inside the
KAZA-TFCA was good, yet few dispersers successfully moved from Zambia's NPs into
neighboring areas. By rendering interactions between movement preferences and
habitat preferences, our movement model substantially outperformed a model that
omitted such interactions.

Ultimately, we show that a simulation-based approach that leverages on
step-selection functions offers a simple yet powerful alternative to traditional
connectivity modeling techniques. Such an approach not only makes fewer
biologically unrealistic assumptions but also permits a more mechanistic
understanding of dispersal and landscape connectivity. Our approach is thus
useful for a variety of applications in ecological, evolutionary, and
conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. Importantly, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal movements, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used models to assess connectivity are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}). Unfortunately, both models rest on assumptions that appear
unsuitable for dispersers, calling for the development of alternative
approaches. One promising alternative is to assess landscape connectivity via
simulated dispersal trajectories generated from individual-based movement models
(IBMMs, \citep{Diniz.2019}).

@@

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
using IBMMs and by converting simulated trajectories into meaningful measures of
connectivity \citep{Diniz.2019}. In contrast to LCPA and CT, IBMMs allow to
explicitly simulate how individuals move across and interact with the
encountered landscape \citep{Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020}, as well as to render potential interactions
between movement behavior and habitat conditions \citep{Avgar.2016}. This
strictly shifts the focus from a structural to a more functional view on
connectivity \citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}.
Furthermore, IBMMs generate movement sequentially, i.e. they generate a series
of steps, so that the temporal dimension of dispersal movements becomes explicit
and allows modeling autocorrelation between consecutive steps
\citep{Diniz.2019}. Finally, simulations from IBMMs do not enforce movement or
connections towards preconceived endpoints, thereby preventing biases arising
from misplaced endpoints. Despite these advantages, a unifying framework to
simulate dispersal and assess connectivity using IBMMs is lacking. Considering
the large number of subjective decisions entailed by IBMMs, an approach that
streamlines and unifies the application of dispersal simulations to assess
connectivity will, however, be critical to safeguard interspecfic comparability
and reproducability.

% Proposed Solution: Three-Step approach
\subsection{Proposed Solution: Three-Step Approach}
Here, we propose and exemplify a simple three-step approach for simulating
dispersal and assessing landscape connectivity (\Cref{GraphicalAbstract}). In
step one, we combine GPS movement data of dispersing individuals with relevant
habitat covariates to fit a mechanistic movement model using integrated step
selection functions (ISSFs, \citealp{Avgar.2016}). ISSFs allow inference on the
study species' habitat kernel (i.e. habitat preferences), its movement kernel
(i.e. movement preferences/capabilities), and potential interactions between the
two \citep{Avgar.2016, Fieberg.2021}. In step two, we simulate individual
dispersal trajectories using the parametrized movement model. Comparable
simulations have already been applied to estimate steady-state utilization
distributions of resident individuals \citep{Potts.2013, Signer.2017} and to
model landscape connectivity, yet disregarding interdependencies between habitat
and movement kernels \citep{Clark.2015, Zeller.2020}. Finally, in step three, we
convert simulated trajectories into three complementary connectivity maps, each
highlighting a different aspect of connectivity; (i) a heatmap revealing areas
that are frequently traversed by dispersers (e.g. \citealp{Hauenstein.2019,
Zeller.2020}), (ii) a betweenness-map delineating dispersal corridors and
bottlenecks (e.g. \citealp{BastilleRousseau.2018}), (iii) and a map of
inter-patch connectivity, depicting the presence and intensity of specific
connections, as well as the average dispersal duration required for the
realization of those connections (e.g. \citealp{Gustafson.1996,
Kanagaraj.2013}).

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed approach (\Cref{GraphicalAbstract})
using GPS movement data collected on dispersing individuals of the endangered
African wild dog (\textit{Lycaon pictus}). The African wild dog is a highly
mobile species whose population persistence heavily relies on the availability
of large, natural or semi-natural landscapes and a sufficient degree of
connectivity among remaining subpopulations. Once common throughout sub-Saharan
Africa, this species has disappeared from much of its historic range, largely
due to human persecution, habitat fragmentation, and disease outbreaks
\citep{Woodroffe.2012}. Wild dogs typically disperse in single-sex coalitions
\citep{McNutt.1996, Behr.2020} and are capable of dispersing several hundred
kilometers \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020}. Although
previous studies have investigated connectivity using LCPA \citep{Hofmann.2021}
or CT \citep{Brennan.2020}, a more comprehensive and mechanistic understanding
of connectivity is missing for this species (but see \citealp{Creel.2020}). With
fewer than 6,000 free-ranging wild dogs remaining in fragmented subpopulations
\citep{Woodroffe.2012}, reliable information on landscape connectivity is
essential for the conservation of this endangered carnivore. Here, we use GPS
data of 16 dispersing wild dogs originating from a free-ranging population in
northern Botswana to parametrize a mechanistic movement model, which we then
employed to simulate 80,000 dispersal trajectories across the landscape of the
world's largest transboundary conservation area, the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA). We anticipated that simulations
based on our three-step approach would overcome several of the above highlighted
conceptual shortcomings of traditional connectivity models and provide a more
comprehensive view on dispersal behavior and landscape connectivity.

@@

\section{Methods}
\subsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA),
an area of approximately 1.3 Mio. km\textsuperscript{2} (\Cref{StudyArea}a and
b). The KAZA-TFCA is the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia, thus hosting
a rich diversity of landscapes, ranging from savannah to grassland and from dry
to moist woodland habitats. In its center lies the Okavango Delta, a dominant
hydro-geographical feature and the world's largest flood-pulsing inland delta.
Large portions of the KAZA-TFCA are formally protected in the form of national
parks (NPs) or other protected areas, yet a considerable portion of the
landscape is still human-dominated (e.g. roads, agricultural sites, and
settlements).

@@

\subsection{Data Collection and Preparation}
\subsubsection{GPS Data}
We collected GPS movement data on 16 dispersing coalitions of African wild dogs
(7 female and 9 male coalitions) between 2011 and 2019 from a free-ranging
population in northern Botswana (details on the data collection can be found in
\cite{Cozzi.2020} and \cite{Hofmann.2021}). To delineate periods of dispersal,
we determined the exact timing of emigration from the natal pack and settlement
in a new territory using direct field observations and through visual inspection
of the net squared displacement (NSD) metric. The NSD metric measures the
squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which we set to the center of the territory of each
dispersers' natal pack. Thus, dispersal was deemed to have started once an
individual left its natal territory and ended once the NSD metric remained
constant, indicating settlement. Because behavior during dispersal is more
pertinent to landscape connectivity than behavior during residence
\citep{Elliot.2014, Abrahms.2017}, we only considered data collected during
dispersal for our analysis. During dispersal, GPS collars recorded a fix every 4
hours and regularly transmitted data over the Iridium satellite system. To
ensure regular time intervals between GPS fixes, we removed any fixes that were
not successfully obtained at the desired 4-hour schedule (allowing for a
tolerance of \( \pm \) 15 minutes). We then converted the fixes (n = 4'169) into
steps, where each step represented the straight-line movement between two
consecutive GPS fixes \citep{Turchin.1998}.

@@

The k-fold cross-validation of the movement model showed that the final model
substantially outperformed a random guess and suggested reliable predictions
(i.e. confidence intervals of \(\bar{r}_{s, realized}\) and \(\bar{r}_{s,
random}\) did not overlap). Moreover, the model correctly assigned high
selection scores to realized steps (\Cref{MovementModel}b), indicating a good
fit between predictions and observations. Compared to the base model
(\(\bar{r}_{s, realized} = -0.55; 95\%-CI = [-0.57, -0.52]\);
\citealp{Hofmann.2021}), the inclusion of several interactions between movement
and habitat covariates significantly improved model performance (\(\bar{r}_{s,
realized} = -0.65; 95\%-CI = [-0.67, -0.64]\).

@@

\begin{figure}
  \includegraphics[width=\textwidth]{99_InterpatchConnectivity.png}
  \caption{Map of inter-patch connectivity in relation to dispersal duration,
  highlighting connections between NPs (dark green). Yellow bubbles represent
  the center of the different NPs and are sized in relation to the number of
  simulated dispersers originating from each park. Black dots represent NPs that
  were smaller than 700 km\textsuperscript{2} and therefore did not serve as
  source areas. Arrows between NPs illustrate between which NPs the simulated
  dispersers successfully moved and the color of each arrow shows the average
  number of steps (4-hourly movements) that were necessary to realize those
  connections. Additionally, the line thickness indicates the relative number of
  dispersers originating from a NP that realized those connections. Note that a
  similar network view could be adopted to investigate connectivity between
  other protected areas need not to be restricted to NPs.}
  \label{InterpatchConnectivity}
\end{figure}

@@

% Short Summary
\subsection{Short Summary}
Here, we proposed a three-step approach to assess landscape connectivity via
simulated dispersal trajectories, and demonstrated its application using
empirical data from a free-ranging population of African wild dogs. In step one,
we used ISSFs to parametrize a fully mechanistic movement model describing how
individuals move through the landscape. In step two, we employed the movement
model to simulate dispersal trajectories across the landscape. In step three, we
translated the simulated trajectories into three complementary connectivity
maps, each emphasizing a different aspect of landscape connectivity (e.g.
frequently traversed areas, critical dispersal corridors and bottlenecks, and
the presence and intensity of functional links between suitable patches).
Importantly, such simulations from ISSFs overcome several conceptual
shortcomings inherent to more traditional connectivity modeling techniques, such
as LCPA and CT.

% Movement Model
\subsection{Movement Model}
Our results on habitat preferences showed that dispersers avoid areas dominated
by humans and covered by water, but select for regions with open grassland in
the vicinity to water bodies. This largely complied with previous studies that
investigated habitat selection by dispersing wild dogs
\citep{DaviesMostert.2012, Masenga.2016, Woodroffe.2019, Oneill.2020,
Hofmann.2021}. However, by also accounting for movement preferences, we were
able to model several additional complexities common to dispersal. For instance,
by including an interaction between turning angle and step length we could
accommodate that dispersers exhibit step lengths that are correlated with
turning angles, meaning turning angles are smaller when individuals move fast.
Although similar autocorrelations could be incorporated by sampling step lengths
and turning angles from copula probability distributions \citep{Hodel.2021a,
Hodel.2021b}, the ISSF framework allowed us to conventiently include
correlations in the movement model. We only considered first order
autocorrelation, i.e. correlation between two consecutive steps, although higher
order autocorrelation is conceivable and may be desirable to model
\citep{Dray.2010, McClintock.2012}. This will, however, require vast amounts of
GPS data that are not interrupted by missing fixes; something that is rarely
achieved in reality \citep{Graves.2006}. The power and flexibility of ISSFs to
model additive effects between habitat and movement covariates
\citep{Avgar.2016, Signer.2017} furthermore allowed us to formally capture that
dispersing wild dogs move slower and more tortuous in areas covered by water,
something that surely was to be expected. Overall, the inclusion of interactions
between habitat and movement covariates in our movement model lead to a
significant improvement in predictive performance compared to an earlier model
that omitted such interactions \citep{Hofmann.2021}.

\subsection{Simulation}
While our approach of simulating dispersal proved usedful to assess landscape
connectivity, it was computationally very costly. Our simulation of 80,000
dispersal trajectories moving 2'000 steps across the KAZA-TFCA required five
days of computation on a regular desktop machine (AMD Ryzen 7 2700X processor
with 8 x 3.6 GHz and 16 logical cores, 64 GB of RAM). The long simulation time
was primarily caused by the massive extent of the study area considered (ca. 1.8
Mio km\textsuperscript{2}) and the large number of trajectories simulated. Most
connectivity studies focus on smaller study areas (e.g. \citealp{Kanagaraj.2013,
Clark.2015, McClure.2016, Abrahms.2017, Zeller.2020}) and will therefore achieve
faster simulation times (given the same spatial resolution across covariates).
We also believe that fewer simulated trajectories will often suffice, as the
relative traversal frequency by simulated trajectories through randomly placed
checkpoints across our study area converged already after 10,500 runs. The
number of required simulations to achieve reliable estimates of connectivity
will, however, vary depending on the structure of the landscape and the
dispersal behavior of the focal species \citep{Gustafson.1996}. For species that
disperse short distances through homogeneous environments, few simulations may
suffice to gauge connectivity, whereas for species that disperse over long
distances through heterogeneous habitats, a large number of simulations will be
required to sufficiently explore the spectrum of possible routes.

### 2022-01-24 Chapter_1/04_Manuscript/Manuscript_Support.tex
updates

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{\textbf{Appendix}\\ A Three Steps Workflow for Simulating Dispersal
Trajectories and Assessing Landscape Connectivity: A Case Study on the
Endangered African Wild Dog}

>>>>

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{\textbf{Appendix}\\ A Thee-Step Approach for Assessing Landscape
Connectivity via Simulated Dispersal: African Wild Dog Case Study}
